---
title: "Titanic Analysis"
author: "Jerry Kiely"
date: "14 February 2015"
output:
  pdf_document: null
  html_document:
    keep_md: yes
    theme: cerulean
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}

    set.seed(1337);

    library(lattice,      warn.conflicts = FALSE, quietly = TRUE);
    library(ggplot2,      warn.conflicts = FALSE, quietly = TRUE);

    library(plyr,         warn.conflicts = FALSE, quietly = TRUE);

    library(survival,     warn.conflicts = FALSE, quietly = TRUE);
    library(splines,      warn.conflicts = FALSE, quietly = TRUE);
    library(parallel,     warn.conflicts = FALSE, quietly = TRUE);
    library(stabs,        warn.conflicts = FALSE, quietly = TRUE);

    library(randomForest, warn.conflicts = FALSE, quietly = TRUE);
    library(gbm,          warn.conflicts = FALSE, quietly = TRUE);
    library(mboost,       warn.conflicts = FALSE, quietly = TRUE);
    library(caret,        warn.conflicts = FALSE, quietly = TRUE);

    source('titanic.R', local = TRUE);

```


## The Introduction

Load the cleaned data:

```{r}

    train <- load_training_data();
    test  <- load_testing_data();

    str(train);

```

##The Analysis

Partition the data into training and cross validation:

```{r}

    partition <- createDataPartition(y = train$Survived, p = 0.75, list = FALSE);
    part_tr   <- train[ partition, ];
    part_cv   <- train[-partition, ];

```

Train the Random Forest model:

```{r, warning=FALSE, message=FALSE}

    model1 <- train(
        Survived ~ ., 
        method     = 'rf', 
        data       = part_tr,
        importance = TRUE,
        trControl  = trainControl(method = 'oob', number = 4)
    );

```

Now train the glm model:

```{r, warning=FALSE, message=FALSE}

    model2 <- train(
        Survived ~ ., 
        method     = 'glm', 
        data       = part_tr,
        family     = binomial
    );

```

Now train the gbm model:

```{r, warning=FALSE, message=FALSE}

    model3 <- train(
        Survived ~ ., 
        method     = 'gbm', 
        data       = part_tr, 
        verbose    = FALSE
    );

```

Look at the importance of the various features / variables of the Random Forest model:

```{r, warning=FALSE, message=FALSE}
 
     importance1 <- varImp(model1);
 
     importance1;
     plot(importance1);
 
```

Look at the importance of the various features / variables of the glm model:

```{r, warning=FALSE, message=FALSE}
 
     importance2 <- varImp(model2);
 
     importance2;
     plot(importance2);
 
```

Look at the importance of the various features / variables of the gbm model:

```{r, warning=FALSE, message=FALSE}
 
     importance3 <- varImp(model3);
 
     importance3;
     plot(importance3);
 
```

Look at the final model of the Random Forest:

```{r, warning=FALSE, message=FALSE}

    model1;
    model1$finalModel

```

Look at the final model of the glm:

```{r, warning=FALSE, message=FALSE}

    model2;
    model2$finalModel

```

Look at the final model of the gbm:

```{r, warning=FALSE, message=FALSE}

    model3;
    model3$finalModel

```

Predict with the training set using the Random Forest model:

```{r, warning=FALSE, message=FALSE}

    predict1_tr <- predict(model1, part_tr);
    cm_tr       <- confusionMatrix(predict1_tr, part_tr$Survived);
    cm_tr;

```

Predict with the cross validation set using the Random Forest model:

```{r, warning=FALSE, message=FALSE}

    predict1_cv <- predict(model1, part_cv);
    cm_cv       <- confusionMatrix(predict1_cv, part_cv$Survived);
    cm_cv;

```

Predict with the training set using the glm model:

```{r, warning=FALSE, message=FALSE}

    predict2_tr <- predict(model2, part_tr);
    cm_tr       <- confusionMatrix(predict2_tr, part_tr$Survived);
    cm_tr;

```

Predict with the cross validation set using the glm model:

```{r, warning=FALSE, message=FALSE}

    predict2_cv <- predict(model2, part_cv);
    cm_cv       <- confusionMatrix(predict2_cv, part_cv$Survived);
    cm_cv;

```

Predict with the training set using the gbm model:

```{r, warning=FALSE, message=FALSE}

    predict3_tr <- predict(model3, part_tr);
    cm_tr       <- confusionMatrix(predict3_tr, part_tr$Survived);
    cm_tr;

```

Predict with the cross validation set using the gbm model:

```{r, warning=FALSE, message=FALSE}

    predict3_cv <- predict(model3, part_cv);
    cm_cv       <- confusionMatrix(predict3_cv, part_cv$Survived);
    cm_cv;

```

Fit a model that includes all predictors:

```{r, warning=FALSE, message=FALSE}

    pred_tr <- data.frame(
        prediction1 = predict1_tr, 
        prediction2 = predict2_tr, 
        prediction3 = predict3_tr, 
        Survived    = part_tr$Survived
    );

    pred_cv <- data.frame(
        prediction1 = predict1_cv, 
        prediction2 = predict2_cv, 
        prediction3 = predict3_cv
    );

    comb_model  <- train(
        Survived ~ ., 
        method = 'gamboost', 
        data   = pred_tr
    );

```

Look at the importance of the various features / variables of the combined model:

```{r, warning=FALSE, message=FALSE}
 
     importance4 <- varImp(comb_model);
 
     importance4;
     plot(importance4);

```

Look at the final model of the combined model:

```{r, warning=FALSE, message=FALSE}

    comb_model;

```

Predict with the training set using the combined model:

```{r, warning=FALSE, message=FALSE}

    predict4_tr <- predict(comb_model, pred_tr);
    cm_tr       <- confusionMatrix(predict4_tr, part_tr$Survived);
    cm_tr;

```

Predict with the cross validation set using the combined model:

```{r, warning=FALSE, message=FALSE}

    predict4_cv <- predict(comb_model, pred_cv);
    cm_cv       <- confusionMatrix(predict4_cv, part_cv$Survived);
    cm_cv;

```


