---
title: "Analyzing Data From A Personal Activity Monitoring Device"
output:
  html_document:
    keep_md: yes
    theme: cerulean
  pdf_document: null
---

## Introduction

It is now possible to collect a large amount of data about personal movement using activity monitoring devices such as a 
Fitbit, Nike Fuelband, or Jawbone Up. These type of devices are part of the "quantified self" movement – a group of 
enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, 
or because they are tech geeks. But these data remain under-utilized both because the raw data are hard to obtain and 
there is a lack of statistical methods and software for processing and interpreting the data. 

This assignment makes use of data from a personal activity monitoring device. This device collects data at 5 minute 
intervals through out the day. The data consists of two months of data from an anonymous individual collected during 
the months of October and November, 2012 and include the number of steps taken in 5 minute intervals each day.


## Loading and preprocessing the data

First I loaded the required libraries, then unzipped the archive (if it wasn't already unzipped), then I load the dataset.

```{r required_libs}
    if (!require('plyr')) {

        install.packages('plyr');
        require('plyr');
    }

    if (!require('reshape2')) {

        install.packages('reshape2');
        require('reshape2');
    }

    if (!file.exists('activity.csv')) {

        unzip('./activity.zip', exdir = '.');
    }


    activity <- read.csv('activity.csv');
```

## What is mean total number of steps taken per day?

To answer this question I used reshape - melt followed by downcast - to get aggregations of steps by day.
Then I plotted a histogram of the dataset, to get an idea of how the steps per day are distributed.

```{r activity_date_hist}
    activity_long <- melt(activity, id.vars = "date", measure.vars = "steps", na.rm = TRUE);
    activity_wide <- dcast(activity_long, date ~ variable, sum);

    hist(activity_wide$steps, 
         col = 'grey', 
         main = 'Total Number Of Steps', 
         xlab = 'Steps'
    );
```

I calculated the mean and median.

```{r mean_median}
    steps_mean   <- mean(activity_wide$steps);
    steps_median <- median(activity_wide$steps);
```

The mean was `r format(steps_mean, scientific = FALSE)` and the median was `r format(steps_median, scientific = FALSE)`.

## What is the average daily activity pattern?

To answer this question I once again used reshape to get aggregations of steps by interval across all days.
Then I plotted a timeseries of the dataset.

```{r plot_interval}
    interval_long <- melt(activity, id.vars = "interval", measure.vars = "steps", na.rm = TRUE);
    interval_wide <- dcast(interval_long, interval ~ variable, mean);

    plot(interval_wide$interval, 
         interval_wide$steps, 
         col = 'blue', 
         main = 'Average Daily Activity Pattern', 
         xlab = '5 Minute Interval', 
         ylab = 'Average Number Of Steps Taken', 
         type = 'l'
    );
```

I calculated the max interval value across all days.

```{r max_steps}
    max_steps    <- max(interval_wide$steps);
    max_interval <- interval_wide[interval_wide$steps == max_steps, 1];
```

The interval with the maximum value was `r max_interval`, with a max value of `r format(max_steps, scientific = FALSE)`.

## Imputing missing values

To begin with, I calculated the total number of missing values (NAs) in the dataset.

```{r sum_nas}
    sum(is.na(activity$steps));
```

Then I assigned the interval mean to the missing values in the dataset, to create a new dataset.

```{r cleaned_data}
    activity_cleaned <- ddply(activity, ~interval, function(item) { 

        item$steps[is.na(item$steps)] <- round(mean(item$steps, na.rm = TRUE));
        return(item);
    });
```

Then to confirm I had provided values for all NAs I calculated the total number of missing values 
in the new dataset.

```{r sum_nas_cleaned}
    sum(is.na(activity_cleaned$steps));
```

Once again I plotted a histogram - this time of the new dataset.

```{r cleaned_hist}
    activity_cleaned_long <- melt(activity_cleaned, id.vars = "date", measure.vars = "steps", na.rm = TRUE);
    activity_cleaned_wide <- dcast(activity_cleaned_long, date ~ variable, sum);

    hist(activity_cleaned_wide$steps, 
         col = 'grey', 
         main = 'Total Number Of Steps', 
         xlab = 'Steps'
    );
```

And I calculated the mean and median of the new dataset.

```{r cleaned_mean_median}
    steps_cleaned_mean   <- mean(activity_cleaned_wide$steps);
    steps_cleaned_median <- median(activity_cleaned_wide$steps);
```

The new mean was `r format(steps_cleaned_mean, scientific = FALSE)` and the median was `r format(steps_cleaned_median, scientific = FALSE)`.
The original mean was `r format(steps_mean, scientific = FALSE)` and the original median was `r format(steps_median, scientific = FALSE)`. So
the process of imputing missing values had a small effect on the statistics of the dataset.

## Are there differences in activity patterns between weekdays and weekends?

To answer this I first added a new column that specified if the row contained weekend or weekday data.

```{r cleaned_weekend}
    activity_cleaned$date    <- as.Date(activity_cleaned$date);
    activity_cleaned$weekend <- ifelse(weekdays(activity_cleaned$date) %in% c('Saturday', 'Sunday'), 'Weekend', 'Weekday');
```

Then I performed the usual aggregation, this time by interval and weekend.

```{r cleaned_weekend_filtered}
    activity_cleaned_long <- melt(activity_cleaned, id.vars = c("interval", "weekend"), measure.vars = "steps", na.rm = TRUE);
    activity_cleaned_wide <- dcast(activity_cleaned_long, interval + weekend ~ variable, mean);

    activity_cleaned_weekend <- subset(activity_cleaned_wide, weekend == 'Weekend');
    activity_cleaned_weekday <- subset(activity_cleaned_wide, weekend == 'Weekday');
```

I then constructed plots for Weekdays and Weekends.

```{r plot_cleaned_weekend}
    par(mfrow = c(2, 1));

    plot(activity_cleaned_weekend$interval, 
         activity_cleaned_weekend$steps, 
         col = 'blue', 
         main = 'Weekend', 
         xlab = 'Interval', 
         ylab = 'Number of Steps', 
         type = 'l'
    );

    plot(activity_cleaned_weekday$interval, 
         activity_cleaned_weekday$steps, 
         col = 'blue', 
         main = 'Weekday', 
         xlab = 'Interval', 
         ylab = 'Number of Steps', 
         type = 'l'
    );
```

I calculated the max, mean, and median values across intervals for weekdays and weekends.

```{r cleaned_seekend_stats}
    weekday_max    <- max(activity_cleaned_weekday$steps);
    weekday_mean   <- mean(activity_cleaned_weekday$steps);
    weekday_median <- median(activity_cleaned_weekday$steps);

    weekend_max    <- max(activity_cleaned_weekend$steps);
    weekend_mean   <- mean(activity_cleaned_weekend$steps);
    weekend_median <- median(activity_cleaned_weekend$steps);

```

The weekday maximum was `r format(weekday_max, scientific = FALSE)` and the weekend maximum 
was `r format(weekend_max, scientific = FALSE)`. The weekday mean was 
`r format(weekday_mean, scientific = FALSE)` and the weekend mean was `r format(weekend_mean, scientific = FALSE)`.
The weekday median was `r format(weekday_median, scientific = FALSE)` and the weekend median 
was `r format(weekend_median, scientific = FALSE)`. There seems to be more activity during the weekend as you would
expect (people work during the week), but both weekend and weekdays follow similiar patterns - most activity 
between 800 and around 2100. 
