---
title: "Predicting The Manner Of Exercise"
author: "Jerry Kiely"
date: "22 January 2015"
output:
  pdf_document: null
  html_document:
    keep_md: yes
    theme: cerulean
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}

    set.seed(12345);

    answers <- c('B', 'A', 'B', 'A', 'A', 'E', 'D', 'B', 'A', 'A', 'B', 'C', 'B', 'A', 'E', 'E', 'A', 'B', 'B', 'B');

    library(lattice);
    library(ggplot2);
    library(caret, warn.conflicts = FALSE, quietly = TRUE);

    write_files = function(x) {

        n = length(x)
        for(i in 1:n) {

            filename = paste0("./results/problem_id_", i, ".txt");
            write.table(x[i], file = filename, quote = FALSE, row.names = FALSE, col.names = FALSE);
        }
    }

```


## The Introduction

This project looks at a HAR dataset:

> Human Activity Recognition - HAR - has emerged as a key research area in the last years 
> and is gaining increasing attention by the pervasive computing research community, 
> especially for the development of context-aware systems.

From the project brief:

> Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect 
> a large amount of data about personal activity relatively inexpensively. These type of 
> devices are part of the quantified self movement – a group of enthusiasts who take 
> measurements about themselves regularly to improve their health, to find patterns in their 
> behavior, or because they are tech geeks. 

Relating to the current task:

> One thing that people regularly do is quantify how much of a particular activity they do, 
> but they rarely quantify how well they do it. In this project, your goal will be to use 
> data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They 
> were asked to perform barbell lifts correctly and incorrectly in 5 different ways.  

We will attempt to predict the manner in which participants performed barbell lifts. We 
will do this by training a classifier with the provided dataset.


## The Data

First we load the data:

```{r}

    train <- read.csv('pml-training.csv', na.strings = c('NA', ''));
    test  <- read.csv('pml-testing.csv',  na.strings = c('NA', ''));

```

Next we clean the data by removing any columns that contain null data:

```{r}

    nas   <- colSums(is.na(train)) > 0;

    names <- colnames(train);
    cols  <- names[!nas];
    train <- train[, cols];

    names <- colnames(test);
    cols  <- names[!nas];
    test  <- test[, cols];

```

Next we remove columns for predictors with nearly zero variance - i.e. columns that are 
almost constant and which are unlikely to affect the outcome:

```{r}

    zvs   <- nearZeroVar(train);

    train <- train[, -zvs];
    test  <- test[, -zvs];

```

Then we ommit the first five columns (which contain index, user, and timestamp 
information):

```{r}

    train <- train[, 6:ncol(train)];
    test  <- test[, 6:ncol(test)];

```

Now we can split the training data into two sets - for training and cross validation:

```{r}

    partition <- createDataPartition(y = train$classe, p = 0.8, list = FALSE);
    part_tr   <- train[partition, ];
    part_cv   <- train[-partition, ];

```


## The Training

Now we go straight into building our model - I choose to use a random forest due to 
the number of predictors, with classe as the outcome, and everything else as 
predictors. 

Note: in truth we don't need to perform cross validation as the random forest does 
this internally - we do so only see how well the model performs in terms of 
estimating out of sample error - anything less than 1% is acceptable:

```{r, cache=TRUE, warning=FALSE, message=FALSE}

    model <- train(
        classe ~ ., 
        method = 'rf', 
        data = part_tr, 
        trControl = trainControl(method = 'oob', number = 4)
    );

```

Now we look at how the model fares against the training data, and the cross validation 
data:

```{r, cache=TRUE, warning=FALSE, message=FALSE}

    predict_tr <- predict(model, part_tr);
    cm_tr      <- confusionMatrix(predict_tr, part_tr$classe);

    predict_cv <- predict(model, part_cv);
    cm_cv      <- confusionMatrix(predict_cv, part_cv$classe);

```

Looking at the confusion matrix for the predictions of the training set:

```{r}

    cm_tr;

```

we see an accuracy of 100%, as expected. Looking at the confusion matrix for the 
predictions of the cross validation set:

```{r}

    cm_cv;

```

we see an accuracy of 
`r format((max(model$results$Accuracy)) * 100, digits = 4, scientific = FALSE)`%. 
From this we can estimate the out of sample error to be 
`r format((1 - max(model$results$Accuracy)) * 100, digits = 4, scientific = FALSE)`% 
which is well within the 1% we set for ourselves. 


## The Testing

Now lets run our model against the test set and see how it fares:

```{r, cache=TRUE, warning=FALSE, message=FALSE}

    predict_te <- predict(model, test);
    cm_te      <- confusionMatrix(predict_te, answers);

```

Looking at the confusion matrix for the predictions of the test set against the 
correct outcomes:

```{r}

    cm_te;

```

we see our model has performed with 100% accuracy.

```{r, echo=FALSE}

    write_files(predict_te);

```
