---
title: "The Effect of Vitamin C on Tooth Growth in Guinea Pigs"
author: "Jerry Kiely"
date: "11 November 2014"
output:
  html_document:
    keep_md: yes
    theme: cerulean
  pdf_document: null
---

```{r echo=FALSE}

    library(ggplot2);
    #library(grid);
    library(datasets);

    data(ToothGrowth);
    ToothGrowth$dose <- as.factor(ToothGrowth$dose);

```


## The Introduction

From the R manual:

> The response is the length of odontoblasts (teeth) in each of 10 guinea pigs at each of three dose levels 
> of Vitamin C (0.5, 1, and 2 mg) with each of two delivery methods (orange juice or ascorbic acid).

We will perform an analysis of this data to see if there are any significant differences in tooth growth for
the different delivery methods across dosages.


## The Data

Lets take a look at the data:

```{r raw_data1, cache=TRUE}

    head(ToothGrowth);

```

and now lets take a look at the means of the data, grouped by dose and supp:

```{r raw_data2, cache=TRUE}

    aggregate(len ~ dose + supp, ToothGrowth, mean);

```

We can see, within dosages, and with the exception of 2.0 dosage, that the means vary. Lets illustrate this 
with some visualizations:

```{r barplots, fig.width=9, fig.height=4.5, cache=TRUE}

    plot <- ggplot(ToothGrowth, aes(x = supp, y = len, fill = supp));
    plot <- plot + geom_boxplot();
    plot <- plot + facet_grid(~ dose);
    plot <- plot + xlab("delivery");
    plot <- plot + ylab("tooth length");
    plot;

```

from looking at the boxplots it's clear that dosage 0.5 and 1.0 have different medians for the different 
deliveries. On the other hand, there doesn't seem to be a difference between the medians of the deliveries 
for dosage 2.0. Lets take a look at the confidence intervals for the different means across dosage and 
delivery:

```{r ci_means, fig.width=9, fig.height=4.5, cache=TRUE}

    conf.int <- function(x) {
        n    = length(x);
        mean = mean(x);
        sd   = sd(x);
        se   = sd / sqrt(n);
        t    = qt(0.975, df = n - 1);
        c(mean = mean, ci_lower = mean - t * se, ci_upper = mean + t * se);
    }

    ci        <- do.call(data.frame, aggregate(len ~ dose + supp, ToothGrowth, conf.int));
    names(ci) <- c('dose', 'supp', 'mean', 'ci_lower', 'ci_upper');

    plot <- ggplot(ci, aes(x = supp, y = mean, colour = supp));
    plot <- plot + geom_point();
    plot <- plot + geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2);
    plot <- plot + facet_grid(~ dose);
    plot <- plot + ggtitle("Confidence Intervals");
    plot <- plot + xlab("delivery");
    plot <- plot + ylab("mean tooth length");
    plot;

```

this shows that the deliveries of both dosage 0.5 and 1.0 have different means - even their confidence 
intervals don't overlap - but the means of the deliveries for dosage 2.0 appear to be the same.


## The Analysis

Lets run some hypothesis tests across the different dosages to see if the means are significantly different. 

\begin{align*}
   H_0 & : \qquad \text{ the true difference in means is equal to zero }
   \\
   H_a & : \qquad \text{ the true difference in means is not equal to zero }
\end{align*}

we assume the samples are independent, and that the subjects are randomly assigned. Also, as the size of 
each sample is less than 10% of the size of the population, we have met the necessary conditions to perform 
our hypothesis tests. First we prepare our data:

```{r test_data, cache=TRUE}

    dose_0.5    <- subset(ToothGrowth, dose == 0.5);
    dose_1.0    <- subset(ToothGrowth, dose == 1.0);
    dose_2.0    <- subset(ToothGrowth, dose == 2.0);

    dose_0.5_VC <- subset(dose_0.5, supp == 'VC');
    dose_1.0_VC <- subset(dose_1.0, supp == 'VC');
    dose_2.0_VC <- subset(dose_2.0, supp == 'VC');

    dose_0.5_OJ <- subset(dose_0.5, supp == 'OJ');
    dose_1.0_OJ <- subset(dose_1.0, supp == 'OJ');
    dose_2.0_OJ <- subset(dose_2.0, supp == 'OJ');

```

then we run our tests. We note that the data isn't paired. Also from looking at the boxplots it looks as if 
the samples have different variances - in fact, when in doubt, we should err on the side of unequal variances.

```{r hypothesis_tests, cache=TRUE}

    dose_0.5_test <- t.test(dose_0.5_OJ$len, dose_0.5_VC$len);
    dose_1.0_test <- t.test(dose_1.0_OJ$len, dose_1.0_VC$len);
    dose_2.0_test <- t.test(dose_2.0_OJ$len, dose_2.0_VC$len);

```

the first test looks at the difference between the two delivery methods across the 0.5 dosage:

```{r hypothesis_test1, cache=TRUE}

    dose_0.5_test;

```

with a p-value of `r dose_0.5_test$p.value` we reject the null hypothesis that the true difference in means 
is equal to zero at the 5% significance level. Looking at the associated confidence interval 
(`r dose_0.5_test$conf.int`) we see it doesn't include 0, which is what we would expect at that p-value. The 
next test looks at the difference between the two delivery methods across the 1.0 dosage:

```{r hypothesis_test2, cache=TRUE}

    dose_1.0_test;

```

with a p-value of `r dose_1.0_test$p.value` we reject the null hypothesis that the true difference in means 
is equal to zero at the 5% significance level. Looking at the associated confidence interval 
(`r dose_1.0_test$conf.int`) we see it doesn't include 0, which is what we would expect at that p-value. The 
next test looks at the difference between the two delivery methods across the 2.0 dosage:

```{r hypothesis_test3, cache=TRUE}

    dose_2.0_test;

```

with a p-value of `r dose_2.0_test$p.value` we fail to reject the null hypothesis that the true difference 
in means is equal to zero at the 5% significance level. Looking at the associated confidence interval 
(`r dose_2.0_test$conf.int`) we see it includes 0, which is what we would expect at that p-value. Lets 
visualize the confidence intervals:

```{r ci_differences, fig.width=9, fig.height=4.5, cache=TRUE}

    dose_0.5_CI <- dose_0.5_test$conf.int;
    dose_1.0_CI <- dose_1.0_test$conf.int;
    dose_2.0_CI <- dose_2.0_test$conf.int;

    dose_0.5_M  <- dose_0.5_test$estimate[1] - dose_0.5_test$estimate[2];
    dose_1.0_M  <- dose_1.0_test$estimate[1] - dose_1.0_test$estimate[2];
    dose_2.0_M  <- dose_2.0_test$estimate[1] - dose_2.0_test$estimate[2];

    stats <- data.frame(
        supp       = factor(c('0', '0', '0')),
        dose       = factor(c('0.5', '1.0', '2.0')),
        difference = c(dose_0.5_M, dose_1.0_M, dose_2.0_M),
        ci_lower   = c(dose_0.5_CI[1], dose_1.0_CI[1], dose_2.0_CI[1]),
        ci_upper   = c(dose_0.5_CI[2], dose_1.0_CI[2], dose_2.0_CI[2])
    );

    plot <- ggplot(stats, aes(x = supp, y = difference, colour = dose));
    plot <- plot + geom_point();
    plot <- plot + geom_errorbar(aes(ymin = ci_lower, ymax = ci_upper), width = 0.2);
    plot <- plot + facet_grid(~ dose);
    plot <- plot + theme(axis.ticks = element_blank(), axis.text.x = element_blank());
    plot <- plot + guides(fill = FALSE);
    plot <- plot + ggtitle("Confidence Intervals");
    plot <- plot + xlab("delivery");
    plot <- plot + ylab("difference in mean tooth length");
    plot;

```


## The Conclusion

It seems clear that for the 0.5 and 1.0 dosages there is a significant difference between the different deliveries 
and the effects on tooth growth, with orange juice being superior to ascorbic acid. There is no significant 
difference between the deliveries for the 2.0 dosage.
